Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-03-17 Write-ups/Manuscript.Rmd
simulation section of WNAR paper

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(here)
library(ROCR)

set.seed(314)
```

@@

We use two metrics to evaluate the out-of-sample predictive performance: integrated squared error (ISE) and Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction accucay of latent continuous function. It is evaluated on the binned grid at midpoints of each unobserved bin. If the entire functional domian has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$.

The second metric, AUC, focuses on evaluation of prediction of the binary outcome. Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid as well and estimated values of latent functions between bins with linear interpolation. Just like ISE, the AUC report is also the average value across the whole sample.

@@

## Result

```{r}
load(here("Data/sim_data.RData"))
load(here("Data/SimOutput_fGFPCA.RData"))

@@

# code prediction value at observed points as NA
for(m in 1:length(pred_lst)){
  pred_lst[[m]]$pred_t200[pred_lst[[m]]$bin <= 200] <- NA
  pred_lst[[m]]$pred_t400[pred_lst[[m]]$bin <= 400] <- NA
  pred_lst[[m]]$pred_t600[pred_lst[[m]]$bin <= 600] <- NA
  pred_lst[[m]]$pred_t800[pred_lst[[m]]$bin <= 800] <- NA
}

@@

```{r}
# true data
df_true <- sim_data[[1]] %>% select(id, sind_inx, eta_i)

# prediction results used for visualization
exp_df <- pred_lst[[1]]

# randomly draw four samples for visualization
rand_id <- sample(exp_df$id, 4)

exp_df %>%
  left_join(df_true %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_i))+
  geom_line(aes(x=bin, y = pred_t200, col = "200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t400, col = "400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t600, col = "600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t800, col = "800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")

@@

```{r}
runtime
# calculate ISE on binned grid
ise_mat <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
colnames(ise_mat) <- c("t200", "t400", "t600", "t800")

for(m in 1:M){
  # true value at midpoints
  true_eta <- sim_data[[m]] %>% filter(sind_inx %in% mid) %>% select(id, sind_inx, eta_i) %>% rename(bin=sind_inx)
  pred_eta <- pred_lst[[m]] %>% select(-eta_hat)
  pred_eta <- pred_eta %>% left_join(true_eta, by = c("id", "bin"))

  ise_m <- c(
    t200 = sum((pred_eta$pred_t200-pred_eta$eta_i)^2, na.rm = T)/N,
    t400 = sum((pred_eta$pred_t400-pred_eta$eta_i)^2, na.rm = T)/N,
    t600 = sum((pred_eta$pred_t600-pred_eta$eta_i)^2, na.rm = T)/N,
    t800 = sum((pred_eta$pred_t800-pred_eta$eta_i)^2, na.rm = T)/N
  )

  ise_mat[m, ] <- ise_m
}
```
```{r}
# calculate AUC on unbinned grid
auc_mat <- matrix(NA, nrow = M, ncol = 4) # row index iteration, col index max observation time
colnames(auc_mat) <- c("t200", "t400", "t600", "t800")

for(m in 1:M){
  # true value as matrix
  Y_mat <- sim_data[[m]]  %>% select(id, sind_inx, Y) %>% rename(bin=sind_inx) %>% 
    pivot_wider(names_from = id, values_from = Y) %>%
    column_to_rownames("bin")

  # calculate AUC
  t_vec <- c(200, 400, 600, 800)
  for(i in seq_along(t_vec)){

    t <- t_vec[i]
    # predicted value
    pred_name <- paste("pred_t", t, sep = "")
    pred_mat <- pred_lst[[m]] %>% select(id, bin, all_of(pred_name)) %>% 
      pivot_wider(names_from = id, values_from = pred_name) %>%
      column_to_rownames("bin")
    pred_full <- sapply(1:N,
                           function(x){approx(x=mid, y=pred_mat[ , x], xout = 1:J)$y})
    pred_full <- exp(pred_full)/(1+exp(pred_full))
    row_id <- complete.cases(pred_full)
    auc_perf <- performance(prediction(pred_full[row_id, ], Y_mat[row_id, ]), "auc")
    auc_mat[m, i] <- mean(unlist(auc_perf@y.values))
  }

auc_mat
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
library(here)
library(ROCR)
library(kableExtra)
library(knitr)
options(knitr.kable.NA = "")

set.seed(314)

load(here("Data/sim_data.RData"))
load(here("Data/SimOutput_fGFPCA.RData"))
load(here("Data/SimOutput_GLMMadaptive.RData"))
```

@@

We use two metrics to evaluate the out-of-sample predictive performance: integrated squared error (ISE) and Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction accucay of latent continuous function. It is evaluated on the binned grid at midpoints of each unobserved bin. If the entire functional domian has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$. The second metric, AUC, focuses on evaluation of prediction of the binary outcome. Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid as well and estimated values of latent functions between bins with linear interpolation.

@@

The average ISE and AUC across all simulation is presented in Table 1, and Figure 1 presents the predicted function curves of four randomly-drawn subjects from the first simulation. The prediction is conditioning on different length of observed track (specifically, with observations up to t = 200, 400, 600, 800 respectively), and evaluation is made on equal-length time windows on the unobserved tracks following the maximum observation time. As Table 1 reveals, the fGFPCA outperformed GLMMadaptive under every scenario, which was expected since GLMMadaptive can only accomodate simple model structures. In addition, fGFPCA also takes less computation time. For one simulated dataset, fGFPCA spent `r round(mean(runtime),2)` minutes on model fitting and out-of-sample prediction, while GLMMadaptive took `r round(mean(runtime_ref),2)` minutes.

Because of the flexibility of our proposed model framework, the accuracy of prediction at specific time points would improve with more observed data. This is revealed in Table 1 as ISE decreases and AUC increases with maximum observed time (left-to-right), also in Figure 1 as predicted curves get closer with longer observed track. However the same tendency is not observed with GLMMadaptive models, as a result of restricted model flexibility. A linear model that fits well to a specific part of latent function can fit very badly to the following parts, especially when the underlying latent function has cyclic patterns. Therefore, more observations do not necessarily make the model more predictive.

```{r}

@@

# for fGFPCA
# code prediction value at observed points as NA
for(m in 1:length(pred_lst)){
  pred_lst[[m]]$pred_t200[pred_lst[[m]]$bin <= 200] <- NA
  pred_lst[[m]]$pred_t400[pred_lst[[m]]$bin <= 400] <- NA
  pred_lst[[m]]$pred_t600[pred_lst[[m]]$bin <= 600] <- NA
  pred_lst[[m]]$pred_t800[pred_lst[[m]]$bin <= 800] <- NA
}

# for GLMM adaptive
# add in observation time index
for(m in 1:length(pred_lst_ref)){
  pred_lst_ref[[m]]$sind_inx <- rep(1:J, N)
}
```

```{r Figure, fig.cap="Predicted track of four randomly selected subjects. The grey dash line indicates the true latent continuous function. The dashed lines indicated predicted latent function tracks, and color indicates different observations time."}
# figure of fGFPCA prediction
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# true data
df_true_sim1 <- sim_data[[1]] %>% select(id, sind_inx, eta_i)

# prediction results used for visualization
exp_df_sim1 <- pred_lst[[1]]

# randomly draw four samples for visualization
rand_id <- sample(exp_df_sim1$id, 4)

exp_df_sim1 %>%
  left_join(df_true_sim1 %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_i, col = "True"))+
  geom_line(aes(x=bin, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Bin", colour= "Observed time")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))
```

```{r ISE_fGFPCA}
# calculate ISE on binned grid
ise_mat <- array(NA, dim = c(5, 4, M)) # dimensions are: predicted time, max obs time, simulation

for(m in 1:M){

  # true value at midpoints
  true_eta <- sim_data[[m]] %>% filter(sind_inx %in% mid) %>% select(id, sind_inx, eta_i) %>% rename(bin=sind_inx)
  pred_eta <- pred_lst[[m]] %>% select(-eta_hat)
  pred_eta <- pred_eta %>% left_join(true_eta, by = c("id", "bin"))

  # split unobserved time into intervals of length 200
  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
  pred_eta$bin_cat <- cut(pred_eta$bin, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)

  # SE at each observed time
  pred_eta <- pred_eta %>% mutate(ise_t200 = (pred_t200-eta_i)^2,
                                  ise_t400 = (pred_t400-eta_i)^2,
                                  ise_t600 = (pred_t600-eta_i)^2,
                                  ise_t800 = (pred_t800-eta_i)^2)

  # subject-average SE
  ise_mat[, , m] <- pred_eta %>% group_by(bin_cat) %>% 
    summarize_at(vars(starts_with("ise")), .fun=function(x)(sum(x)/N)) %>%select(-bin_cat) %>% as.matrix()
}

ave_ise_mat <- apply(ise_mat, MARGIN = c(1, 2), mean)[-1, ]
colnames(ave_ise_mat) <- c("200", "400", "600", "800")
rownames(ave_ise_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

@@

```{r, ISE_GLMMadaptive}
# calculate ISE on binned grid
ise_mat_ref <- array(NA, dim = c(5, 4, M)) # dimensions are: predicted time, max obs time, simulation

for(m in 1:M){

  # true value at midpoints
  pred_eta <- pred_lst_ref[[m]] %>% filter(sind_inx %in% mid)

  # split unobserved time into intervals of length 200
  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
  pred_eta$bin_cat <- cut(pred_eta$sind_inx, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)

  # SE at each observed time
  pred_eta <- pred_eta %>% mutate(ise_t200 = (pred_t200-eta_i)^2,
                                  ise_t400 = (pred_t400-eta_i)^2,
                                  ise_t600 = (pred_t600-eta_i)^2,
                                  ise_t800 = (pred_t800-eta_i)^2)

  # subject-average SE
  ise_mat_ref[, , m] <- pred_eta %>% group_by(bin_cat) %>% 
    summarize_at(vars(starts_with("ise")), .fun=function(x)(sum(x)/N)) %>%select(-bin_cat) %>% as.matrix()

ave_ise_mat_ref <- apply(ise_mat_ref, MARGIN = c(1, 2), mean)[-1, ]
colnames(ave_ise_mat_ref) <- c("200", "400", "600", "800")
rownames(ave_ise_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

# cbind(ave_ise_mat, ave_ise_mat_ref)
```

```{r AUC_fGFPCA, cache=TRUE}
# maximum observations time
t_vec <- c(200, 400, 600, 800)

# calculate AUC on unbinned grid
auc_mat <- array(NA, dim = c(4, 4, M)) # row index iteration, col index max observation time

for(m in 1:M){

  Y_df <- sim_data[[m]]
  eta_pred <- pred_lst[[m]]

  # use interpolation to extend prediction to the unbinned grid
  for(i in 1:N){
   Y_df$pred_200[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t200[eta_pred$id==i], xout = 1:J)$y
   Y_df$pred_400[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t400[eta_pred$id==i], xout = 1:J)$y
   Y_df$pred_600[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t600[eta_pred$id==i], xout = 1:J)$y
   Y_df$pred_800[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t800[eta_pred$id==i], xout = 1:J)$y
  }

  # split unobserved time into intervals of length 200
  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
  Y_df$bin_cat <- cut(Y_df$sind_inx, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
  Y_df_lst <- split(Y_df, f = Y_df$bin_cat)
  Y_df_lst <- Y_df_lst[-1]

  for(j in seq_along(t_vec)){

    t <- t_vec[j]
    pred_name <- paste("pred_", t, sep = "")

    # calculate AUC
    auc_lst <- lapply(Y_df_lst[j:4], 
                       function(x){
                         row_id <- !is.na(x[, pred_name])
                         this_eta <- x[row_id, pred_name]
                         this_Y <- x[row_id, "Y"]
                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
                         return(auc_pref@y.values)
                })

    auc_vec <- c(rep(NA, 4-length(auc_lst)), unlist(auc_lst))
    auc_mat[ , j, m] <- auc_vec

}}

ave_auc_mat <- apply(auc_mat, MARGIN = c(1, 2), mean)
colnames(ave_auc_mat) <- c("200", "400", "600", "800")
rownames(ave_auc_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

@@

```{r AUC_GLMMadaptive}
# calculate AUC on unbinned grid
auc_mat_ref <- array(NA, dim = c(4, 4, M)) # row index iteration, col index max observation time

for(m in 1:M){

  Y_df <- sim_data[[m]]
  eta_pred <- pred_lst_ref[[m]]

  Y_df <- Y_df %>% left_join(eta_pred %>% select(id, sind_inx, starts_with("pred_")), by = c("id", "sind_inx"))

  # split unobserved time into intervals of length 200
  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
  Y_df$bin_cat <- cut(Y_df$sind_inx, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
  Y_df_lst <- split(Y_df, f = Y_df$bin_cat)
  Y_df_lst <- Y_df_lst[-1]

  for(j in seq_along(t_vec)){

    t <- t_vec[j]
    pred_name <- paste("pred_t", t, sep = "")

    # calculate AUC
    auc_lst <- lapply(Y_df_lst[j:4], 
                       function(x){
                         row_id <- !is.na(x[, pred_name])
                         this_eta <- x[row_id, pred_name]
                         this_Y <- x[row_id, "Y"]
                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
                         return(auc_pref@y.values)
                })

    auc_vec <- c(rep(NA, 4-length(auc_lst)), unlist(auc_lst))
    auc_mat_ref[ , j, m] <- auc_vec

  }
}

ave_auc_mat_ref <- apply(auc_mat_ref, MARGIN = c(1, 2), mean)
colnames(ave_auc_mat_ref) <- c("200", "400", "600", "800")
rownames(ave_auc_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

```{r format table}
ise_tb <- cbind(ave_ise_mat, ave_ise_mat_ref)
auc_tb <- cbind(ave_auc_mat, ave_auc_mat_ref)

rbind(ise_tb, auc_tb) %>%
  kable(booktabs = T, digit=2,
        caption = "Predictive performance of fGFPCA and GLMM adaptive on the simulated datasets. ISE and AUC are average values across all 50 simulations.",
        align = "lcccccccc") %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
  add_header_above(c(" " = 1, "Maximum observed time" = 8)) %>% 
  group_rows(index = c("Prediction time window" = 8)) %>%
  group_rows(index = c("ISE" = 4, "AUC" = 4)) 
```

### 2023-03-17 Write-ups/Manuscript.tex
solve AUC error
simulation of GLMM adaptvie
simulation section of WNAR paper

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage[]{mathpazo}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}\doublespacing
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis},
  pdfauthor={Ying Jin; Andrew Leroux},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{March 14, 2023}

@@

We use two metrics to evaluate the out-of-sample predictive performance:
integrated squared error (ISE) and
Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction
accucay of latent continuous function. It is evaluated on the binned
grid at midpoints of each unobserved bin. If the entire functional
domian has S bins, but we have observations up to the mth bin, then ISE
is defined as
\(\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).

The second metric, AUC, focuses on evaluation of prediction of the
binary outcome. Since the binary outcomes are generated on the original,
un-binned grid, we evaluated AUC on this grid as well and estimated
values of latent functions between bins with linear interpolation. Just
like ISE, the AUC report is also the average value across the whole
sample.

@@

\hypertarget{result}{%
\subsection{Result}\label{result}}

\includegraphics{Manuscript_files/figure-latex/unnamed-chunk-2-1.pdf}

\begin{verbatim}
## [1] 5.033985 5.134139 5.195642 5.152902 5.082466
\end{verbatim}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage[]{mathpazo}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}\doublespacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis},
  pdfauthor={Ying Jin; Andrew Leroux},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{March 17, 2023}

@@

We use two metrics to evaluate the out-of-sample predictive performance:
integrated squared error (ISE) and
Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction
accucay of latent continuous function. It is evaluated on the binned
grid at midpoints of each unobserved bin. If the entire functional
domian has S bins, but we have observations up to the mth bin, then ISE
is defined as
\(\frac{1}{N}\sum_{i=1}^N\sum_{s=m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2\).
The second metric, AUC, focuses on evaluation of prediction of the
binary outcome. Since the binary outcomes are generated on the original,
un-binned grid, we evaluated AUC on this grid as well and estimated
values of latent functions between bins with linear interpolation.

@@

The average ISE and AUC across all simulation is presented in Table 1,
and Figure 1 presents the predicted function curves of four
randomly-drawn subjects from the first simulation. The prediction is
conditioning on different length of observed track (specifically, with
observations up to t = 200, 400, 600, 800 respectively), and evaluation
is made on equal-length time windows on the unobserved tracks following
the maximum observation time. As Table 1 reveals, the fGFPCA
outperformed GLMMadaptive under every scenario, which was expected since
GLMMadaptive can only accomodate simple model structures. In addition,
fGFPCA also takes less computation time. For one simulated dataset,
fGFPCA spent 4.94 minutes on model fitting and out-of-sample prediction,
while GLMMadaptive took 5.11 minutes.

Because of the flexibility of our proposed model framework, the accuracy
of prediction at specific time points would improve with more observed
data. This is revealed in Table 1 as ISE decreases and AUC increases
with maximum observed time (left-to-right), also in Figure 1 as
predicted curves get closer with longer observed track. However the same
tendency is not observed with GLMMadaptive models, as a result of
restricted model flexibility. A linear model that fits well to a
specific part of latent function can fit very badly to the following
parts, especially when the underlying latent function has cyclic
patterns. Therefore, more observations do not necessarily make the model
more predictive.

\begin{figure}
\centering
\includegraphics{Manuscript_files/figure-latex/Figure-1.pdf}
\caption{Predicted track of four randomly selected subjects. The grey
dash line indicates the true latent continuous function. The dashed
lines indicated predicted latent function tracks, and color indicates
different observations time.}
\end{figure}

\begin{table}

\caption{\label{tab:format table}Predictive performance of fGFPCA and GLMM adaptive on the simulated datasets. ISE and AUC are average values across all 50 simulations.}
\centering
\begin{tabular}[t]{llccccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Maximum observed time} \\
\cmidrule(l{3pt}r{3pt}){2-9}
\multicolumn{1}{c}{ } & \multicolumn{4}{c}{fGFPCA} & \multicolumn{4}{c}{GLMMadaptive} \\
\cmidrule(l{3pt}r{3pt}){2-5} \cmidrule(l{3pt}r{3pt}){6-9}
  & 200 & 400 & 600 & 800 & 200 & 400 & 600 & 800\\
\midrule
\addlinespace[0.3em]
\multicolumn{9}{l}{\textbf{Prediction time window}}\\
\addlinespace[0.3em]
\multicolumn{9}{l}{\textbf{ISE}}\\
\hspace{1em}\hspace{1em}(200, 400] & 15.29 &  &  &  & 38.20 &  &  & \\
\hspace{1em}\hspace{1em}(400, 600] & 18.68 & 8.53 &  &  & 28.59 & 26.46 &  & \\
\hspace{1em}\hspace{1em}(600, 800] & 22.53 & 5.51 & 1.65 &  & 31.32 & 27.97 & 27.40 & \\
\hspace{1em}\hspace{1em}(800, 1000] & 11.34 & 8.72 & 1.90 & 1.30 & 55.19 & 46.89 & 58.57 & 59.25\\
\addlinespace[0.3em]
\multicolumn{9}{l}{\textbf{AUC}}\\
\hspace{1em}\hspace{1em}(200, 400] & 0.74 &  &  &  & 0.59 &  &  & \\
\hspace{1em}\hspace{1em}(400, 600] & 0.66 & 0.73 &  &  & 0.52 & 0.59 &  & \\
\hspace{1em}\hspace{1em}(600, 800] & 0.71 & 0.79 & 0.80 &  & 0.67 & 0.70 & 0.69 & \\
\hspace{1em}\hspace{1em}(800, 1000] & 0.74 & 0.75 & 0.78 & 0.78 & 0.52 & 0.56 & 0.53 & 0.57\\
\bottomrule
\end{tabular}
\end{table}
